---
title: "ares: Fast MARS in roadrunner"
author: "Jack Trametta"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ares: Fast MARS in roadrunner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

`ares()` (shipped in the `roadrunner` package) fits Multivariate Adaptive
Regression Splines (MARS) models — a flexible nonparametric regression
technique introduced by Friedman (1991). The MARS forward pass adds basis
functions of the form `max(0, ±(x - knot))` (called *hinges*) one pair at a
time, each scaled by a parent term to allow interactions. A backward pass
then prunes terms by minimizing the GCV criterion.

`ares()` mirrors the API of [`earth`](https://cran.r-project.org/package=earth)
on the gaussian-only core and is designed for numerical parity with `earth`
while taking advantage of multi-core CPUs via `RcppParallel`.

## A small example: `mtcars`

```{r}
library(roadrunner)
fit <- ares(mpg ~ ., data = mtcars, nthreads = 2)
print(fit)
```

The fit returns a list with the standard MARS components.

```{r}
str(fit, max.level = 1)
```

`predict()` produces fitted values on training data or new data.

```{r}
head(predict(fit))
head(predict(fit, mtcars[1:5, ]))
```

## Comparison with `earth`

When the `earth` package is available, we can directly compare fits.

```{r}
if (requireNamespace("earth", quietly = TRUE)) {
  set.seed(20260509)
  n <- 200; p <- 5
  x <- matrix(runif(n * p), n, p)
  y <- 10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 0.5)^2 +
       10 * x[, 4] + 5 * x[, 5] + rnorm(n)
  fa <- ares(x, y, degree = 2, nthreads = 2)
  fe <- earth::earth(x, y, degree = 2)
  cat("ares  RSS:", round(fa$rss, 3), " GCV:", round(fa$gcv, 3),
      " nterms:", length(fa$selected.terms), "\n")
  cat("earth RSS:", round(fe$rss, 3), " GCV:", round(fe$gcv, 3),
      " nterms:", length(fe$selected.terms), "\n")
  cat("relative RSS difference:",
      round(abs(fa$rss - fe$rss) / fe$rss, 4), "\n")
}
```

The relative RSS difference is typically a fraction of a percent at n >= 1000 —
the two implementations agree numerically even when they pick slightly different
knot sets.

## Threading determinism

`ares` is parallel-deterministic by construction: changing `nthreads` does not
change the output.

```{r}
set.seed(42)
n <- 150; p <- 5
x <- matrix(runif(n * p), n, p)
y <- 5 * pmax(0, x[, 1] - 0.5) * pmax(0, x[, 2] - 0.3) + rnorm(n)
f1 <- ares(x, y, degree = 2, nthreads = 1)
f2 <- ares(x, y, degree = 2, nthreads = 2)
cat("identical selected.terms:", isTRUE(all.equal(f1$selected.terms, f2$selected.terms)), "\n")
cat("max coef diff:           ", max(abs(f1$coefficients - f2$coefficients)), "\n")
```

## Performance notes

`ares` v0.0.0.9000 is currently **slower in absolute wall-clock than `earth`** —
this is a documented limitation and the headline target for v0.1.
Internally, `ares` uses an O(n) per-knot scoring loop where `earth` uses
Friedman's O(1)-per-knot Givens fast-LS update. The Givens implementation is on
the roadmap.

What `ares` v0.0.0.9000 already gets right:

- Numerical correctness — fits track `earth` to ~1% RSS across the smoke grid.
- Threading determinism — bit-identical fits regardless of thread count.
- Parallel scaling — 2-thread is consistently ~1.5–1.8× faster than 1-thread
  (the `RcppParallel` worker is doing real work; the constant factor lives
  inside the inner loop, not in the threading layer).

See the `inst/sims/` directory for the Monte Carlo benchmark scripts that
generated the headline numbers in `README.md`.

## References

Friedman, J. H. (1991). Multivariate Adaptive Regression Splines.
*Annals of Statistics* 19(1):1–67.
