---
title: "ares: Fast MARS in roadrunner"
author: "Jack T. Rametta"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ares: Fast MARS in roadrunner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

`ares()` fits Multivariate Adaptive Regression Splines (MARS) models, a
flexible nonparametric regression technique introduced by Friedman
(1991). The forward pass adds basis functions of the form
`max(0, ±(x - knot))` one pair at a time, each scaled by a parent term
to allow interactions. A backward pass prunes terms by minimising the
GCV criterion (or a K-fold CV criterion when `pmethod = "cv"` or
`nfold > 0`).

`ares()` mirrors the API of
[`earth`](https://cran.r-project.org/package=earth) on the gaussian core
and adds: hyperparameter autotune, K-fold CV pruning, row-bootstrap
bagging, observation weights, binomial / poisson / gamma GLM families on
the selected basis, and residual-variance prediction intervals.

## A small example: `mtcars`

```{r}
library(roadrunner)
fit <- ares(mpg ~ ., data = mtcars, nthreads = 2)
print(fit)
```

```{r}
head(predict(fit))
head(predict(fit, mtcars[1:5, ]))
```

## Comparison with `earth`

```{r}
if (requireNamespace("earth", quietly = TRUE)) {
  set.seed(20260509)
  n <- 200; p <- 5
  x <- matrix(runif(n * p), n, p)
  y <- 10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 0.5)^2 +
       10 * x[, 4] + 5 * x[, 5] + rnorm(n)
  fa <- ares(x, y, degree = 2, nthreads = 2)
  fe <- earth::earth(x, y, degree = 2)
  cat("ares  RSS:", round(fa$rss, 3), " GCV:", round(fa$gcv, 3),
      " nterms:", length(fa$selected.terms), "\n")
  cat("earth RSS:", round(fe$rss, 3), " GCV:", round(fe$gcv, 3),
      " nterms:", length(fe$selected.terms), "\n")
  cat("relative RSS difference:",
      round(abs(fa$rss - fe$rss) / fe$rss, 4), "\n")
}
```

The two implementations agree numerically to within a fraction of a
percent on RSS even when they pick slightly different knot sets.

## Threading determinism

`ares` is parallel-deterministic by construction: changing `nthreads`
does not change the output bit-for-bit.

```{r}
set.seed(42)
n <- 150; p <- 5
x <- matrix(runif(n * p), n, p)
y <- 5 * pmax(0, x[, 1] - 0.5) * pmax(0, x[, 2] - 0.3) + rnorm(n)
f1 <- ares(x, y, degree = 2, nthreads = 1)
f2 <- ares(x, y, degree = 2, nthreads = 2)
identical(f1$selected.terms, f2$selected.terms)
max(abs(f1$coefficients - f2$coefficients))
```

## Hyperparameter autotune

`autotune = TRUE` runs an inner K-fold CV grid search over
`(degree, penalty, nk, fast.k)` and refits the winner on the full data.
A warm-start phase fits a small subsample first and short-circuits if
the best-per-degree gap is decisive.

```{r}
set.seed(7)
fit_at <- ares(mpg ~ ., data = mtcars, autotune = TRUE,
               autotune.speed = "fast", seed.cv = 1L, nthreads = 2)
fit_at$autotune$degree
fit_at$autotune$penalty
fit_at$autotune$nk
```

## Classification via `family = "binomial"`

Forward + backward run as gaussian; the selected basis is then refit
with `stats::glm.fit(family = binomial())`. `predict()` returns
probabilities by default, or the linear predictor via `type = "link"`.

```{r}
set.seed(1)
n <- 400
x <- matrix(rnorm(n * 4), n, 4)
p <- plogis(0.5 * x[, 1] + 1.2 * x[, 2] * (x[, 2] > 0) - 0.4 * x[, 3])
y <- rbinom(n, 1, p)
fb <- ares(x, y, family = "binomial", degree = 2, nthreads = 2)
range(predict(fb))
```

## Prediction intervals

`varmod = "const"` or `varmod = "lm"` (gaussian only) stores a residual
variance model at fit time. `predict(interval = "pint")` then returns a
matrix with `c("fit", "lwr", "upr")` columns.

```{r}
set.seed(0)
n <- 500
x <- matrix(runif(n * 3), n, 3)
y <- 3 * x[, 1] + 2 * sin(2 * pi * x[, 2]) + rnorm(n, sd = 0.5)
fp <- ares(x, y, degree = 2, varmod = "const", nthreads = 2)
head(predict(fp, x[1:3, ], interval = "pint"))
```

## References

Friedman, J. H. (1991). Multivariate Adaptive Regression Splines.
*Annals of Statistics* 19(1):1–67.
